---
title: Clearer, better, faster, reproducible bacterial genus delineation procedure
affiliations:
  - id: uka
    name: University Hospital of RWTH Aachen
    department: Institute of Medical Microbiology
    group: Functional Microbiome Research Group
    country: Germany
    url: https://ror.org/02gm5zw39
authors:
  - name: Charlie Pauvert
    corresponding: true
    orcid: 0000-0001-9832-2507
    email: cpauvert@ukaachen.de
    affiliations:
      - ref: uka
  - name: Thomas C.A. Hitch
    orcid: 0000-0003-2244-7412
    affiliations:
      - ref: uka
  - name: Thomas Clavel
    orcid: 0000-0002-7229-5595
    affiliations:
      - ref: uka
license: "CC BY"
bibliography: references.bib
---

```{r setup}
library(tidyverse)
library(targets)
```


# Introduction

```{r cpu_hours}
overall_stats <- tar_read(family_metadata) %>%
  select(Family,benchmark_type, n_genomes,CPU_hours) %>%
  mutate(
    n_comparisons=n_genomes*(n_genomes-1),
    n_comparisons=if_else(benchmark_type=="full",
                          n_comparisons*10, n_comparisons)
                          ) %>%
  summarise(
    n = sum(n_comparisons),
    total_hours = sum(CPU_hours) %>% round(),
    total_years = as.period(total_hours, unit = "hours") %>% as.numeric("years")
  ) %>% mutate(across(n:total_years, ~prettyNum(.x,big.mark = " ", digits=2)))
```

Still, this study needed `r overall_stats$n` pairwise comparisons for a total of `r overall_stats$total_hours` CPU-Hours (`r overall_stats$total_years` in years).

# Methods

## GTDB bacterial proteins sequences and taxonomy as gold standard

Publicly-available microbial proteins sequences from shortlisted bacteria were used for our comparison.
The Genome Taxonomy DataBase (GTDB) is a well-maintained and metadata rich microbial resource that provides curated taxonomy along genomes and genome-derived proteins sequences [@parksGTDBOngoingCensus2022].
We devised inclusion criteria to span a wide breadth of the phylogenetic tree while maintaining achievable comparisons with the time, human and computing resources available.

Any bacterial entry available in the GTDB (r214) was included provided (1) the bacteria has a valid name according to the LPSN, (2) the entry is a representative genome, (3) the bacteria belongs to a family with at least two genuses, and (5) it belongs to a genus with at least ten genomes.

Since version r214, GTDB provides proteins sequences FASTA file (`.faa`) as compressed files (`.gz`).
While we support the decision to save storage and network load, we had to add an extraction step before use to ensure a fair comparison as only BLAST cannot natively process compressed FASTA files.

We used r214.0 and are aware that r214.1 corrected the taxonomy assignation of the genome `GB_GCA_902406375.1` to *Collinsella sp002232035*.
This change did not impact our study as the genome in question is not part of the shortlist.

The bacterial phylogenetic tree of the shortlisted genomes was constructed by trimming the bacterial tree released by GTDB using custom Python scripts and the ETE3 library [@huerta-cepasETEReconstructionAnalysis2016].


Expected pairwise comparisons between a query sequence ($Q$) and a subject sequence ($S$) were defined by banning self-comparisons ($Q \neq S$) and considering reciprocal comparisons ($Q\text{-}S$ and $S\text{-}Q$) only within the same family to prevent exponential explosion.


## Definition of Percentage of Conserved Proteins (POCP)

The percentage of conserved proteins (POCP) between two genomes $Q$ and $S$ is defined as:

$$
POCP = \dfrac{C_{QS} + C_{SQ}}{T_Q + T_S} \times 100\%
$$ {#eq-pocp}

where $C_{QS}$ represent the conserved number of proteins from $Q$ when aligned to $S$ and conversely $C_{SQ}$ represent the conserved number of proteins from $S$ when aligned to $Q$, and $T_Q + T_S$ represent the total number of proteins in the two genomes being compared, respectively [adapted from @qinProposedGenusBoundary2014].
The range of POCP is theoretically $\left[0-100\%\right]$.
Conserved proteins are defined as protein sequences matches from the query with an e-value < $10^{−5}$, a sequence identity > 40%, and an aligned region > 50% of the query protein sequence length.


However, proteins sequences from the query can match multiple subject sequences in the case of duplicated genes.
While briefly mentioned in the original paper that "The number of conserved proteins in each genome of strains being compared was slightly different because of the existence of duplicate genes (paralogs)" [@qinProposedGenusBoundary2014, p. 2211], the expected influence on the POCP values is lacking.
In our experience, this results in POCP values above the theoretical upper bound of 100%.
Therefore, we defined the POCP with unique matches (POCPu) between two genomes $Q$ and $S$ as:

$$
POCPu = \dfrac{C_{uQS} + C_{uSQ}}{T_Q + T_S} \times 100\%
$$ {#eq-pocpu}

where $C_{uQS}$ represent the conserved number of proteins from *the unique matches of* $Q$ when aligned to $S$ and conversely $C_{uSQ}$ represent the conserved number of proteins from *the unique matches of* $S$ when aligned to $Q$, and $T_Q + T_S$ represent the total number of proteins in the two genomes being compared, respectively.
Thus the range of POCP is actually $\left[0-\infty\right[$ and the range of POCPu is $\left[0-100\%\right]$.
Note that the assumption regarding unique matches is made implicitly when using some POCP implementation [e.g., @holzerHoelzerPocp2020; @holzerPOCPnfAutomaticNextflow2024; @linSilentGeneBiopyBiopy2021].

## Approaches for proteins sequence alignment benchmark

Many-versus-many proteins sequences comparisons are required to assess the number of conserved proteins.
While @qinProposedGenusBoundary2014 provided limited guidance on how to implement the computation of POCP, it is assumed that the BLASTP algorithm is the standard.
It is the approach implemented in Protologger [@hitch_automated_2021].
We first use the BLASTP approach [@camachoBLASTArchitectureApplications2009] using the parameters fitting @qinProposedGenusBoundary2014 procedure @tbl-tools-parameters.
We also considered a slightly altered BLASTP approach, named BLASTPDB where BLAST databases are first built for the two genomes considered @tbl-tools-parameters.
This allow parallel alignments on multiple CPU, which is not possible with BLASTP.
We then included two tools that are usually used as faster BLASTP alternatives: DIAMOND [@buchfinkSensitiveProteinAlignments2021] and MMseqs2 [@steineggerMMseqs2EnablesSensitive2017].
The former is being used in @holzerPOCPnfAutomaticNextflow2024 while the latter is used in EzAAI [@kimIntroducingEzAAIPipeline2021].
Similarly to BLASTPDB, these approaches require that a protein database is built for each genome before performing the alignment @tbl-tools-parameters.
DIAMOND and MMseqs2 were both used with four different sensitivity thresholds @tbl-tools-parameters on a recent comparisons [@buchfinkSensitiveProteinAlignments2021].

```{r tbl-tools-parameters, tab.cap="List of the ten approaches and associated parameters for the many-versus-many proteins alignments tools used in the benchmark. The recommended approach is indicated in bold."}
tibble::tribble(
  ~Name, ~Parameters,
  "BLASTP","--evalue 0.00001 --qcov_hsp_perc 50.0",
  "BLASTPDB","--evalue 0.00001 --qcov_hsp_perc 50.0",
  "DIAMOND_FAST","--evalue 0.00001 --query-cover 50.0 --fast",
  "DIAMOND_SENSITIVE","--evalue 0.00001 --query-cover 50.0 --sensitive",
  "DIAMOND_VERYSENSITIVE","--evalue 0.00001 --query-cover 50.0 --very-sensitive",
  "DIAMOND_ULTRASENSITIVE","--evalue 0.00001 --query-cover 50.0 --ultra-sensitive",
  "MMSEQS2_S1DOT0","--e-profile 0.00001 --cov-mode 1 -c 0.50 -s 1.0",
  "MMSEQS2_S2DOT5","--e-profile 0.00001 --cov-mode 1 -c 0.50 -s 2.5",
  "MMSEQS2_S6DOT0","--e-profile 0.00001 --cov-mode 1 -c 0.50 -s 6.0",
  "MMSEQS2_S7DOT5","--e-profile 0.00001 --cov-mode 1 -c 0.50 -s 7.5"
) %>% 
  mutate(
    Name = if_else(Name=="DIAMOND_VERYSENSITIVE", "**DIAMOND_VERYSENSITIVE**", Name),
    `Needs database?` = if_else(Name == "BLASTP", "No", "Yes"),
    Parameters = glue::glue("`{p}`", p=Parameters)
    ) %>%
  rename("Approach name"="Name") %>% 
  select(`Approach name`, `Needs database?`, Parameters) %>% knitr::kable()
```

All proteins matches were filtered to keep only matches with > 40 % identity to all the query sequences matches for POCP ($C_{QS}$ and $C_{SQ}$ in @eq-pocp) and only unique query sequence matches for POCPu ($C_{QS}$ and $C_{SQ}$ in @eq-pocpu).
Note that the filtering was adapted to the method as the range of percentage of identity in MMseqs2 is $[0-1]$ and $[0-100]$ for BLAST and DIAMOND.
The total number of proteins per genomes ($T_Q$ and $T_S$ in @eq-pocp and @eq-pocpu) was computed using seqkit stats v2.2.0 [@shenSeqKitCrossPlatformUltrafast2016].

Linear regressions were used in `r R.version.string` to fit the expected POCP (or POCPu) values obtained via the legacy BLASTP approach against the others approaches considered.
The coefficient of determination R^2^ of the linear regression is used as an interpretable and bounded goodness-of-fit measure between the expected and measured values instead of others errors measurement [@chiccoCoefficientDeterminationRsquared2021].
We did not rely on the adjusted coefficient of determination as the linear regressions had only one predictor.


## Metrics to evaluate genus delineation

We computed classification metrics with a positive event defined as "both genomes belong to the same genus".
Thus, for a pair of bacterial genomes with a POCP (or POCPu) > 50%, the pair is a True Positive ($TP$) if the pair belong to the same genus, or if not to False Positive ($FP$).
Conversely, for a pair of bacterial genomes with a POCP (or POCPu) $\leq$ 50%, the pair is a False Negative ($FN$) if the pair belong to the same genus, or if not to True Negative ($TN$).
We use the Sensitivity ($\frac{TP}{TP+FN}$), Specificity ($\frac{TN}{TN+FP}$), and the False-discovery rate ($\frac{FP}{FP+TP}$) to assess the classification performance of both POCP and POCPu.


## Workflow implementation

Automatic protein sequences download, data preprocessing, many-versus-many protein alignments, as well as POCP computation and delineation metrics calculations were run using the workflow tool nextflow [@ditommasoNextflowEnablesReproducible2017], as well as components of nf-core [@ewelsNfcoreFrameworkCommunitycurated2020].
On top of a workflow manager, tools used in the benchmark are shipped within Docker container [@merkelDockerLightweightLinux2014]  or bioconda [@thebiocondateamBiocondaSustainableComprehensive2018a] environments to ensure reproducibility, scalability and ease future extensions of the benchmark.

Nextflow natively keep track of the time, CPU, memory and disk usage of each process in an execution trace log file that we used to evaluate computing resources utilization. Process duration is available as walltime and realtime, the CPU usage is reported as a percentage of usage of a unique CPU, meaning multi-threaded processes will have a value higher than 100%.

Statistical analyses, data crunching and visualisation were orchestrated with the R workflow manager targets v.1.7.0 [@landauTargetsPackageDynamic2021].

# References

::: {#refs}
:::
